---
title: "Sample Overlap Reconciliation: DivCo + AMP-AD 1.0"
subtitle: "Mapping genotyped sample IDs to cohort metadata"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

<style>
body  { font-size: 16px; line-height: 1.6; }
h1    { font-size: 24px; font-weight: bold; }
h2    { font-size: 20px; font-weight: bold; }
h3    { font-size: 18px; }
pre, code { font-size: 13px; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 4,
  fig.align  = "center"
)

# global plot theme so all ggplots inherit readable font sizes
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

# Genotype Data 

The genotyped data is the combined **DivCo + AMP-AD 1.0** set (syn67158725). The
sample IDs in the VCF are a mix of identifiers from several contributing studies,
and the goal here is to **generate combined individual, assay, and biospecimen metadata**
by matching it to one of the source metadata files.

```{r libraries, echo = FALSE}
library(dplyr)
library(ggVennDiagram)   
library(ggplot2)
```

```{r helper, echo = FALSE}
# Small helper to draw a 2-set Venn for "genotyped samples" vs. "a metadata source"
venn_two <- function(genotyped, metadata_ids, metadata_label,
                     title = NULL) {
  sets <- list(
    `Genotyped sample IDs` = unique(as.character(genotyped)),
    metadata_ids        = unique(as.character(metadata_ids))
  )
  names(sets)[2] <- metadata_label
  ggVennDiagram(sets, label = "count", label_alpha = 0) +
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
    labs(title = title) +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    theme(plot.margin = unit(c(0.5, 15, 1, 5), "cm"))
}
```

```{r load-sample-list, echo = FALSE}
# Genotyped sample IDs extracted from the filtered VCF
sample_list <- read.table("~/data/processed/sample_ids.txt")$V1
sample_list <- as.character(sample_list)
```
 The full genotyped list contains `r length(sample_list)` samples.

# DivCo metadata

The Diverse Cohorts (DivCo) study provides three metadata tables — individual,
assay (WGS), and biospecimen.

AMP-AD_DiverseCohorts_assay_WGS_metadata.csv is the DivCo metadata file (syn51757644) to which sample list were matched using `specimenID`
AMP-AD_DiverseCohorts_individual_metadata_harmonized.csv (syn73713769) contains individual level data for DivCo cohort
AMP-AD_DiverseCohorts_biospecimen_metadata.csv (syn51757645) contains individual to biospecimen mapping for DivCo cohort 

```{r divco-wgs, echo = FALSE}
AMP_AD_DivCo_wgs_df <- read.table(
  "~/data/metadata/AMP-AD_DivCo/AMP-AD_DiverseCohorts_assay_WGS_metadata.csv",
  sep = ",", header = TRUE
)

# 743 of the DivCo samples are present (of 762 DivCo specimens total)
divco_overlap <- sample_list[sample_list %in% AMP_AD_DivCo_wgs_df$specimenID]

```

The WGS metadata file contains `r length(unique(AMP_AD_DivCo_wgs_df$specimenID))` samples of 
which `r length(divco_overlap)` samples are in DivCo + AMP-AD 1.0 jointcall VCFs

```{r divco-wgs-venn, echo = FALSE}
venn_two(sample_list, AMP_AD_DivCo_wgs_df$specimenID,
         "DivCo specimenID",
         title = "VCF samples vs. DivCo WGS assay metadata")
```
```{r divco-individual, echo = FALSE}
# load individual and biospecimen metadata
AMP_AD_DivCo_individual_df <- read.table(
  "~/data/metadata/harmonized_metadata/AMP-AD_DiverseCohorts_individual_metadata_harmonized.csv",
  sep = ",", header = TRUE,
  quote = "",          # disable quote parsing
  fill = TRUE,         # fill missing fields
  comment.char = ""    # ignore # as comment character
)

AMP_AD_DivCo_biospec_df <- read.table(
  '~/data/metadata/AMP-AD_DivCo/AMP-AD_DiverseCohorts_biospecimen_metadata.csv', 
  sep = ',', header = T)

# Subset specimen df to the ones present in sample list 
AMP_AD_DivCo_biospec_df <- AMP_AD_DivCo_biospec_df[AMP_AD_DivCo_biospec_df$specimenID %in% divco_overlap,]
AMP_AD_DivCo_individual_df <- AMP_AD_DivCo_individual_df[AMP_AD_DivCo_individual_df$individualID %in% AMP_AD_DivCo_biospec_df$individualID, ]
```
# AMP-AD 1.0 metadata

The remaining genotyped samples come from the AMP-AD 1.0 studies (ROSMAP, MSBB,
Mayo). Each is checked against its WGS assay metadata.

### ROSMAP

ROSMAP_assay_wholeGenomeSeq_metadata.csv (syn21314542) is the Rosmap WGS metadata file to which sample list were matched using `specimenID`
ROSMAP_clinical_harmonized.csv (syn73713768) is the Rosmap individual metadata file
ROSMAP_biospecimen_metadata.csv (syn21323366) is the Rosmap biospecimen metadata file 

```{r rosmap_wgs, echo=FALSE}
rosmap_wgs_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/ROSMAP_assay_wholeGenomeSeq_metadata.csv",
  sep = ",", header = TRUE
)
rosmap_overlap <- sample_list[sample_list %in% rosmap_wgs_df$specimenID]
#length(rosmap_overlap)   # 12 overlap
```

The WGS metadata file contains `r length(unique(rosmap_wgs_df$specimenID))` samples of 
which `r length(rosmap_overlap)` samples are in DivCo + AMP-AD 1.0 jointcall VCFs

```{r rosmap-venn, echo = FALSE}
venn_two(sample_list, rosmap_wgs_df$specimenID,
         "ROSMAP specimenID",
         title = "VCF samples vs. ROSMAP WGS assay metadata")
``` 
```{r rosmap_metadata, echo=FALSE}
# load individual and biospecimen metadata
rosmap_individual_df <- read.table(
  "~/data/metadata/harmonized_metadata/ROSMAP_clinical_harmonized.csv",
  sep = ",", header = TRUE,
  quote = "",          # disable quote parsing
  fill = TRUE,         # fill missing fields
  comment.char = ""    # ignore # as comment character
)
rosmap_bio_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/ROSMAP_biospecimen_metadata.csv",
  sep = ",", header = TRUE
)

```
### MSBB
MSBB_assay_wholeGenomeSeq_metadata.csv (syn22360825) is the MSBB WGS metadata file to which sample list were matched using `specimenID` 
MSBB_individual_metadata_harmonized.csv (syn73713767) is the MSBB individual metadata file
MSBB_biospecimen_metadata.csv (syn21893059) is the MSBB biospecimen metadata file 

```{r msbb, echo = FALSE}
msbb_wgs_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/MSBB_assay_wholeGenomeSeq_metadata.csv",
  sep = ",", header = TRUE
)
msbb_overlap <- sample_list[sample_list %in% as.character(msbb_wgs_df$specimenID)]
```
The WGS metadata file contains `r length(unique(msbb_wgs_df$specimenID))` samples of 
which `r length(msbb_overlap)` samples are in DivCo + AMP-AD 1.0 jointcall VCFs
```{r msbb-venn, echo = FALSE}
venn_two(sample_list, msbb_wgs_df$specimenID,
         "MSBB specimenID",
         title = "VCF samples vs. MSBB WGS assay metadata")
```
```{r msbb_all, echo = FALSE}
# load individual and biospecimen metadata
msbb_individual_df <- read.table(
  "~/data/metadata/harmonized_metadata/MSBB_individual_metadata_harmonized.csv",
  sep = ",", header = TRUE,
  quote = "",          # disable quote parsing
  fill = TRUE,         # fill missing fields
  comment.char = ""    # ignore # as comment character
)
msbb_bio_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/MSBB_biospecimen_metadata.csv",
  sep = ",", header = TRUE
)
```
### MayoRNAseq

MayoRNAseq_assay_wholeGenomeSeq_metadata.csv (syn23481994) is the Mayo WGS metadata file to which sample list were matched using `specimenID` 
MayoRNAseq_individual_metadata.csv (syn73713766)  is the Mayo individual metadata file
MayoRNAseq_biospecimen_metadata.csv (syn20827192) is the Mayo biospecimen metadata file 
```{r mayo, echo = FALSE}
mayo_wgs_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/MayoRNAseq_assay_wholeGenomeSeq_metadata.csv",
  sep = ",", header = TRUE
)
mayo_overlap <- sample_list[sample_list %in% as.character(mayo_wgs_df$specimenID)]
```
The WGS metadata file contains `r length(unique(mayo_wgs_df$specimenID))` samples of 
which `r length(mayo_overlap)` samples are in DivCo + AMP-AD 1.0 jointcall VCFs

```{r mayo-venn, echo = FALSE}
venn_two(sample_list, mayo_wgs_df$specimenID,
         "Mayo specimenID",
         title = "Genotyped samples vs. Mayo WGS assaymetadata")
```
```{r mayo_all, echo=FALSE}
# for MayoRNAseq, specimenID and individualID are the same
mayo_individual_df <- read.table(
  "~/data/metadata/harmonized_metadata/MayoRNAseq_individual_metadata_harmonized.csv",
  sep = ",", header = TRUE,
  quote = "",          # disable quote parsing
  fill = TRUE,         # fill missing fields
  comment.char = ""    # ignore # as comment character
)

mayo_bio_df <- read.table(
  "~/data/metadata/AMP-AD_1.0/MayoRNAseq_biospecimen_metadata.csv",
  sep = ",", header = TRUE
)
```

# Combined overlap (first pass)

Accumulate every sample matched so far across DivCo and the three AMP-AD 1.0 studies,
then see how many genotyped samples remain unaccounted for.

```{r combined-first, echo = FALSE}
combined_overlap <- c(mayo_overlap, msbb_overlap, rosmap_overlap, divco_overlap)
# samples still unmatched
remaining <- setdiff(sample_list, combined_overlap)
```
At this stage, `r length(unique(combined_overlap))` samples are accounted for and `r length(setdiff(sample_list, combined_overlap))`
samples are still unmatched — these are resolved below via the DivCo-to-AMP-AD 1.0 ID linking file.

```{r combined-first-venn, echo = FALSE}
venn_two(sample_list, combined_overlap,
         "DivCO + AMP AD 1.0 specimenID",
         title = "VCF samples vs. DivCO + AMP AD 1.0 WGS assay metadata")
```
### Resolving ID-scheme mismatches (DivCo <-> AMP-AD 1.0)
DivCo2_AMPAD1_IDlink2.csv (syn75361064) is a linking files for the n=276 AMP-AD1.0 samples in the DivCo + AMP 1.0 joint called sample set,
listing the individualID and AMP-AD1.0 WGS specimenID (to link to AMP-AD1.0 metadata)

```{r idlink, echo = FALSE}
divco_ampad1_idlink <- read.table(
  "~/data/metadata/DivCo2_AMPAD1_IDlink2.csv",
  sep = ",", header = TRUE
)
mayo_divco_ampad1_idlink   <- divco_ampad1_idlink[divco_ampad1_idlink$study_WGS_AMPAD1 == "MAYO_AMPAD1", ]
rosmap_divco_ampad1_idlink <- divco_ampad1_idlink[divco_ampad1_idlink$study_WGS_AMPAD1 == "ROSMAP_AMPAD1", ]
```
### Mayo ID linking
```{r idlink-mayo, echo = FALSE}
# restrict to linked individuals that are in our genotyped list
mayo_divco_ampad1_idlink <- mayo_divco_ampad1_idlink[
  mayo_divco_ampad1_idlink$individualID %in% sample_list, ]

```
The link file contains `r length(mayo_divco_ampad1_idlink$individualID)` samples from MayoRNAseq cohort
of which `r length(intersect(combined_overlap, mayo_divco_ampad1_idlink$specimenID))` have been accounted for.
These samples have identical individualID and specimenID and have multiple assay types ("wholeGenomeSeq" or "snpArray") and 
specimenIDSource ("MayoBrainBank", "BannerSun", or "SNPs" ) associated with them in MayoRNAseq_biospecimen_metadata.csv

### ROSMAP ID linking

```{r idlink-rosmap, echo = FALSE}
rosmap_divco_ampad1_idlink <- rosmap_divco_ampad1_idlink[
  rosmap_divco_ampad1_idlink$individualID %in% sample_list, ]

# where specimenID already equals individualID in our list
rosmap_divco_ampad1_common <- rosmap_divco_ampad1_idlink[
  rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1 %in% sample_list, ]
#setdiff(rosmap_divco_ampad1_common$specimenID_WGS_AMPAD1,
        #rosmap_divco_ampad1_common$individualID)


# the ROSMAP entries whose specimenID is NOT in the list -> match via individualID instead
rosmap_divco_ampad1_no_overlap <- rosmap_divco_ampad1_idlink[
  !(rosmap_divco_ampad1_idlink$specimenID %in% sample_list), ]

rosmap_ind_sample_mismatch <- sample_list[sample_list %in% rosmap_divco_ampad1_no_overlap$individualID]

```
The link file contains `r length(rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1)` samples from ROSMAP cohort
of which `r length(intersect(combined_overlap, rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1))` have been accounted for.
These IDs had identical specimenID and individualID in ROSMAP_biospecimen_metadata.csv across assays and tissue types. They also have identical 
specimenID_WGS_AMPAD1 and individualID in the link file.
The remaining `r length(rosmap_divco_ampad1_no_overlap$specimenID_WGS_AMPAD1)` samples from the ROSMAP cohort in the link file
have different specimenID and individualID in ROSMAP_biospecimen_metadata.csv (i.e. one specimenID for each tissue type). But the individualIDs were used 
for joint call sample labeling, so the overlap with specimenID fails for them. 
```{r idlink-rosmap-venn, echo = FALSE}
venn_two(sample_list, c(rosmap_divco_ampad1_no_overlap$specimenID_WGS_AMPAD1, rosmap_divco_ampad1_common$specimenID_WGS_AMPAD1),
         "ROSMAP individualID",
         title = "Genotyped samples vs. ROSMAP from link file")
```

## Final reconciliation

Add the ROSMAP donors that matched only via `individualID` (their `specimenID` was not
in the genotyped list), then confirm how many samples remain unaccounted for.

```{r final, echo = FALSE}
combined_overlap <- c(combined_overlap, rosmap_divco_ampad1_no_overlap$individualID)

remaining_final <- setdiff(sample_list, combined_overlap)
#length(remaining_final)
```

```{r final-venn, echo = FALSE}
venn_two(sample_list, combined_overlap,
         "Matched (final)",
         title = "Genotyped samples vs. all matched samples (final)")
```

```{r final-report, echo=FALSE, results='asis'}
n_total   <- length(sample_list)
n_matched <- length(intersect(sample_list, combined_overlap))
n_left    <- length(remaining_final)
cat(sprintf(
  "Of **%d** genotyped samples, **%d** are matched to a metadata source and **%d** remain unaccounted for.\n\n",
  n_total, n_matched, n_left
))
if (n_left > 0) {
  cat("Unmatched sample IDs:\n\n")
  cat(paste0("- `", remaining_final, "`"), sep = "\n")
}
```

# Filter data and create a combined metadata file 

```{r WGS_combined_metadata, echo = FALSE}
#Subset each WGS file to samples from sample list and create combined WGS assay metadata
AMP_AD_DivCo_wgs_df <- AMP_AD_DivCo_wgs_df[AMP_AD_DivCo_wgs_df$specimenID %in% sample_list, ]
#Save table 
write.table(AMP_AD_DivCo_wgs_df, '~/data/processed/metadata/DivCO/DivCo_assay_WGS_metadata.csv', sep = ',')
mayo_wgs_df <- mayo_wgs_df[as.character(mayo_wgs_df$specimenID) %in% mayo_overlap, ]
write.table(mayo_wgs_df, '~/data/processed/metadata/Mayo/Mayo_assay_WGS_metadata.csv', sep = ',')

#for rosmap there are 12 samples with identical individual and biospecimenIDs and 170 without.
#for all 182, will subset the rosmap wgs data using the biospecimenID from the rosmap subset of the link file
#for the 170, will replace the specimenID with individualID in the biospecimen and wgs metadata 
rosmap_wgs_df <- rosmap_wgs_df[rosmap_wgs_df$specimenID %in% rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1, ] #dimension is 182 
rosmap_wgs_df$specimenID_AMPAD_1.0 <- rosmap_wgs_df$specimenID
rosmap_wgs_df$specimenID <- rosmap_divco_ampad1_idlink$individualID[
  match(rosmap_wgs_df$specimenID, rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1)
]
write.table(rosmap_wgs_df, '~/data/processed/metadata/Rosmap/Rosmap_assay_WGS_metadata.csv', sep = ',')
#Check if the colnames in the wgs assay df are the same across 
#setdiff(colnames(rosmap_wgs_df), colnames(mayo_wgs_df)) #exact same colnames for these two 
#setdiff(colnames(mayo_wgs_df), colnames(AMP_AD_DivCo_wgs_df)) #only two columns overlap between AMP AD 1.0 and DivCO naively 
#Based on content and column names the following were uniformly named: 
#totalReads and Total.Reads; read lenth and Mean.Read.Length; 
#added site "Mayo" and "Rush" for samples for which metadata was retreived from MayoRNAseq and Rosmap metadata files.   
#added a new column "WGS metdata source file" - syn51757644 - DivCO; syn23481994 - mayo; syn21314542 - rosmap

rosmap_wgs_clean <- rosmap_wgs_df |>
  select(specimenID, GQN, totalReads, readLength) |>
  mutate(Site = "Rush",
         WGS_metadata_source = "syn21314542")

mayo_wgs_clean <- mayo_wgs_df |>
  select(specimenID, GQN, totalReads, readLength) |>
  mutate(Site = "Mayo",
         WGS_metadata_source = "syn23481994",
         specimenID = as.character(specimenID))

divco_wgs_clean <- AMP_AD_DivCo_wgs_df |>
  rename(totalReads = Total.Reads,
         readLength = Mean.Read.Length) |>
  select(specimenID, GQN, totalReads, readLength, Site) |>
  mutate(WGS_metadata_source = "syn51757644")

combined_wgs_df <- bind_rows(
  mutate(rosmap_wgs_clean, across(everything(), as.character)),
  mutate(mayo_wgs_clean,   across(everything(), as.character)),
  mutate(divco_wgs_clean,  across(everything(), as.character))
)
write.table(combined_wgs_df, '~/data/processed/metadata/Combined/DivCo_AMPAD_1.0_assay_WGS_metadata.csv', sep = ',')
```
```{r biospecimen_combined_metadata, echo = FALSE}
#Subset each biospecimen file to samples from sample list and create combined biospecimen metadata
AMP_AD_DivCo_biospec_df <- AMP_AD_DivCo_biospec_df[AMP_AD_DivCo_biospec_df$specimenID %in% sample_list, ]
#Save table 
write.table(AMP_AD_DivCo_biospec_df, '~/data/processed/metadata/DivCO/DivCo_biospecimen_metadata.csv', sep = ',')
mayo_bio_df <- mayo_bio_df[as.character(mayo_bio_df$specimenID) %in% mayo_overlap, ]
write.table(mayo_bio_df, '~/data/processed/metadata/Mayo/Mayo_biospecimen_metadata.csv', sep = ',')
rosmap_bio_df <- rosmap_bio_df[rosmap_bio_df$individualID %in% c(rosmap_divco_ampad1_idlink$individualID), ] 
write.table(rosmap_bio_df, '~/data/processed/metadata/Rosmap/Rosmap_biospecimen_metadata.csv', sep = ',')
#for the combined metadata, only subsetting to wholeGenomeSeq for mayo and rosmap and resetting specimenID to individualID for rosmap
rosmap_bio_df <- rosmap_bio_df[rosmap_bio_df$specimenID %in% rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1 & rosmap_bio_df$assay == 'wholeGenomeSeq', ]
rosmap_bio_df$specimen_ID_AMPAD_1.0 <- rosmap_bio_df$individualID
rosmap_bio_df$specimenID <- rosmap_divco_ampad1_idlink$individualID[
  match(rosmap_bio_df$specimenID, rosmap_divco_ampad1_idlink$specimenID_WGS_AMPAD1)
]
mayo_bio_df <- mayo_bio_df[mayo_bio_df$assay == 'wholeGenomeSeq', ]
mayo_bio_df$specimen_ID_AMPAD_1.0 <- mayo_bio_df$specimenID
AMP_AD_DivCo_biospec_df <- AMP_AD_DivCo_biospec_df[AMP_AD_DivCo_biospec_df$assay == "wholeGenomeSeq", ]
AMP_AD_DivCo_biospec_df$specimen_ID_AMPAD_1.0 <- "NA"
#intersect(colnames(rosmap_bio_df), colnames(mayo_bio_df))
#intersect(colnames(mayo_bio_df), colnames(AMP_AD_DivCo_biospec_df))
#colnames(AMP_AD_DivCo_biospec_df)
common_cols <- c("individualID", "specimenID", "specimen_ID_AMPAD_1.0", "organ", "tissue", 
                 "BrodmannArea", "sampleStatus", "nucleicAcidSource", "assay")
  
rosmap_bio_clean <- rosmap_bio_df |>
  select(all_of(common_cols)) |>
  mutate(across(everything(), as.character),
         biospecimen_metadata_source = "syn21323366")

mayo_bio_clean <- mayo_bio_df |>
  select(all_of(common_cols)) |>
  mutate(across(everything(), as.character),
         biospecimen_metadata_source = "syn20827192")

divco_bio_clean <- AMP_AD_DivCo_biospec_df |>
  select(all_of(common_cols)) |>
  mutate(across(everything(), as.character),
         biospecimen_metadata_source = "syn51757645")

combined_bio_df <- bind_rows(rosmap_bio_clean, mayo_bio_clean, divco_bio_clean)
write.table(combined_bio_df,
            '~/data/processed/metadata/Combined/DivCo_AMPAD_1.0_biospecimen_metadata.csv',
            sep = ',', row.names = FALSE)

```
```{r Individual_combined_metadata, echo = FALSE}
#Subset each individual file to samples from sample list and create combined WGS assay metadata
AMP_AD_DivCo_individual_df <- AMP_AD_DivCo_individual_df[AMP_AD_DivCo_individual_df$individualID %in% AMP_AD_DivCo_biospec_df$individualID, ]
write.table(AMP_AD_DivCo_individual_df, '~/data/processed/metadata/DivCO/DivCo_individual_metadata.csv', sep = ',')

mayo_individual_df <- mayo_individual_df[as.character(mayo_individual_df$individualID) %in% mayo_bio_df$individualID, ]
write.table(mayo_individual_df, '~/data/processed/metadata/Mayo/Mayo_individual_metadata.csv', sep = ',')

rosmap_individual_df <- rosmap_individual_df[rosmap_individual_df$individualID %in% rosmap_divco_ampad1_idlink$individualID, ]
write.table(rosmap_individual_df, '~/data/processed/metadata/Rosmap/Rosmap_individual_metadata.csv', sep = ',')

# Common columns across all three cohorts
common_cols <- c("individualID", "dataContributionGroup", "cohort", "species",
                 "sex", "race", "isHispanic", "ageDeath", "PMI",
                 "apoeGenotype", "apoe4Status", "amyCerad", "amyAny", 
                 "amyThal", "amyA", "Braak", "bScore",
                 "individual_metadata_source")

rosmap_individual_clean <- rosmap_individual_df |>
  mutate(across(everything(), as.character),
         individual_metadata_source = "syn73713768")

mayo_individual_clean <- mayo_individual_df |>
  mutate(across(everything(), as.character),
         individual_metadata_source = "syn73713766")

divco_individual_clean <- AMP_AD_DivCo_individual_df |>
  mutate(across(everything(), as.character),
         individual_metadata_source = "syn73713769")

combined_individual_df <- bind_rows(
  rosmap_individual_clean |> select(all_of(common_cols)),
  mayo_individual_clean   |> select(all_of(common_cols)),
  divco_individual_clean  |> select(all_of(common_cols))
)

write.table(combined_individual_df,
            '~/data/processed/metadata/Combined/DivCo_AMPAD_1.0_individual_metadata.csv',
            sep = ',', row.names = FALSE)
```
```{r print_hist, echo = FALSE}
combined_wgs_df |>
  count(Site, WGS_metadata_source) |>
  mutate(label = paste0(Site, "\n(", WGS_metadata_source, ")")) |>
  ggplot(aes(x = reorder(label, n), y = n, fill = Site)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("Rush" = "#4E79A7", "Mayo" = "#B07AA1",
                                setNames("#F28E2B", unique(combined_wgs_df$Site[
                                  !combined_wgs_df$Site %in% c("Rush","Mayo")][1])))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "WGS samples by metadata source",
       x = NULL, y = "N samples") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold")) +
        theme(plot.margin = unit(c(1, 15, 1, 1), "cm")) 

```
Metadata from each cohort was filtered and combined into unified metadata tables for WGS assay, 
biospecimen, and individual metadata.

**ROSMAP** requires special handling across all three tables: of the `r nrow(rosmap_wgs_df)` ROSMAP 
samples, 12 have identical `individualID` and `specimenID`, while the remaining 170 use different IDs 
across the DivCo and AMP-AD 1.0 datasets. For all 182 samples, the AMP-AD 1.0 `specimenID` is retained 
in a new column (`specimenID_AMPAD_1.0`) and replaced with the DivCo-compatible `individualID` using 
the ID link file. Additionally, ROSMAP uses non-standard column names for several clinical variables 
(`msex` → `sex`, `age_death` → `ageDeath`, `pmi` → `PMI`, `apoe_genotype` → `apoeGenotype`, 
`braaksc` → `Braak`) which were harmonized prior to combining.

For the **WGS assay metadata**, `Total.Reads` → `totalReads` and `Mean.Read.Length` → `readLength` were 
harmonized across cohorts. Site labels (`Rush`, `Mayo`) were added for ROSMAP and Mayo samples as DivCo 
already contains a `Site` column. The final combined table contains `r nrow(combined_wgs_df)` samples 
across `r length(unique(combined_wgs_df$Site))` sites.

For the **biospecimen metadata**, the following columns were retained across all three cohorts: 
`individualID`, `specimenID`, `organ`, `tissue`, `BrodmannArea`, `sampleStatus`, `nucleicAcidSource`, 
and `assay`. All tables were subset to `wholeGenomeSeq` assay only prior to combining. The final 
combined table contains `r nrow(combined_bio_df)` samples.

For the **individual metadata**, the harmonized metadata from syn73713763 was used. Column names were 
consistent across all three cohorts, requiring no renaming. The following columns were retained across 
all cohorts: `individualID`, `dataContributionGroup`, `cohort`, `species`, `sex`, `race`, `isHispanic`, 
`ageDeath`, `PMI`, `apoeGenotype`, `apoe4Status`, `amyCerad`, `amyAny`, `amyThal`, `amyA`, `Braak`, 
and `bScore`. Cohort-specific columns (e.g. `cogdx`, `dcfdx_lv` for ROSMAP; `diagnosis`, `brainWeight` 
for Mayo; `mayoDx`, `ADoutcome` for DivCo) were excluded from the combined table but retained in the 
cohort-specific files. The final combined table contains `r nrow(combined_individual_df)` individuals across `r length(unique(combined_individual_df$cohort))` 
cohorts.

A `*_metadata_source` column was added to each combined table to track the originating Synapse file 
per cohort. All combined tables are saved to `~/data/processed/metadata/Combined/`.


