---
title: "NYGC QC and syn51757644 (AMP Div Co WGS Metadata) check"
subtitle: "Mapping samples between files and accounting for all samples"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

<style>
body  { font-size: 16px; line-height: 1.6; }
h1    { font-size: 24px; font-weight: bold; }
h2    { font-size: 20px; font-weight: bold; }
h3    { font-size: 18px; }
pre, code { font-size: 13px; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 4,
  fig.align  = "center"
)
# global plot theme so all ggplots inherit readable font sizes
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

## Check the overlap between the files 

WGS Assay metadata file = syn51757644 
NYGC QC file = AMP_AD_WGS_sequencing_QC_NYGC.xlsx
Mapping file between NYGC QC and WGS = DivCo_SampleMapping.csv
Sample IDs = Column names from vcf files in syn2580853


```{r libraries, echo = FALSE}
library(dplyr)
library(ggVennDiagram)   
library(ggplot2)
library(xlsx)
library(stringr)
```
```{r load data, echo = FALSE}

wgs_assay_df_old <- read.table('/home/ec2-user/data/metadata/AMP-AD_DivCo/AMP-AD_DiverseCohorts_assay_WGS_metadata.csv', sep = ',', header = T)
wgs_assay_df <- read.csv('/home/ec2-user/data/processed/metadata/DivCO/DivCo_assay_WGS_metadata.csv')

wgs_qc_df <- read.xlsx('/home/ec2-user/data/metadata/AMP-AD_DivCO-WGS/AMP_AD_WGS_sequencing_QC_NYGC.xlsx', sheetIndex = 1)
qc_assay_map <- read.table('/home/ec2-user/data/metadata/AMP-AD_DivCO-WGS/DivCo_SampleMapping.csv', sep = ',', header = T)
qc_assay_map <- qc_assay_map %>% 
                mutate('wgs_specimenID' = case_when(grepl('DLPFC', final_specimenid) ~ paste0(final_specimenid, '_WGS'),
                                                      TRUE ~ final_specimenid))
sample_ids <- read.table('/home/ec2-user/data/processed/sample_ids.txt')
sample_ids <- sample_ids$V1
dropped_samples <- read.table('/home/ec2-user/data/metadata/AMP-AD_DivCO-WGS/dropped_samples.tsv', sep = '\t', header = T)


intersect(dropped_samples$Sample.ID, wgs_assay_df$specimenID)

length(setdiff(wgs_assay_df$specimenID, wgs_qc_df$wgs_specimenID))

length(setdiff(wgs_qc_df$Sample, qc_assay_map$filename.sample))

wgs_qc_df[wgs_qc_df$X..Sequence.Contamination > 2, ]

wgs_qc_df$wgs_specimenID <- qc_assay_map$'wgs_specimenID'[match(wgs_qc_df$Sample, qc_assay_map$filename.sample)]



not_in_map <- setdiff(wgs_qc_df$Sample, qc_assay_map$filename.sample)

dim(wgs_assay_df[wgs_assay_df$Percent.Sequence.Contamination > 2, 'specimenID'])

length(intersect(wgs_assay_df$specimenID, sample_ids))

intersect(setdiff(wgs_assay_df$specimenID, sample_ids), wgs_qc_df$Sample)

intersect(setdiff(wgs_assay_df$specimenID, sample_ids), qc_assay_map$'wgs_specimenID')


wgs_assay_df[wgs_assay_df$specimenID %in% setdiff(wgs_assay_df$specimenID, sample_ids)
, ]





setdiff(filtered_divco_wgs$specimenID, sample_ids)

dim(filtered_divco_wgs[filtered_divco_wgs$Percent.Sequence.Contamination > 2, ])

not_in_map[grepl('DLPFC',not_in_map)] <- paste0(str_replace(not_in_map[grepl('DLPFC',not_in_map)],'-', '_'),'_WGS')

which(not_in_map %in% wgs_assay_df$specimenID)

wgs_assay_df[wgs_assay_df$specimenID %in% not_in_map, ]


wgs_qc_df[wgs_qc_df$Sample %in% setdiff(not_in_map, contam_samples), ]


x <- (setdiff(c(qc_assay_map$wgs_specimenID, dropped_samples$Sample.ID), wgs_assay_df$specimenID))

setdiff(not_in_map, dropped_samples$Sample.ID) == setdiff(not_in_map, contam_samples)
not_in_map

contam_samples <- wgs_qc_df[wgs_qc_df$X..Sequence.Contamination > 2, 'Sample']
setdiff(contam_samples, dropped_samples$Sample.ID)

setdiff(dropped_samples$Sample.ID, x)
#Venn diagram showing overlap between map file and NYGC QC file
ggVennDiagram(list(`Map File "filename.sample"` = qc_assay_map$filename.sample,
                   `NYGC QC "Sample"` = wgs_qc_df$Sample,
                   `Dropped Samples` = dropped_samples$Sample.ID),
              label = "count", label_alpha = 0) +
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
    labs(title = "Map file and NYGC QC file") +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    theme(plot.margin = unit(c(0.5, 15, 1, 5), "cm"))

#Venn diagram showing overlap between map file and WGS assay file 
ggVennDiagram(list(`Map File "specimenid"` = qc_assay_map$final_specimenid,
                   `WGS Assay "specimenID"` = wgs_assay_df$specimenID),
              label = "count", label_alpha = 0) +
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
    labs(title = "Map file and WGS assay metadata file") +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    theme(plot.margin = unit(c(0.5, 15, 1, 5), "cm"))


#Venn diagram showing overlap between map file and WGS assay file 
ggVennDiagram(list(`Map File "wgs_specimenID"` = qc_assay_map$'wgs_specimenID',
                   `WGS Assay "specimenID"` = wgs_assay_df$specimenID),
              label = "count", label_alpha = 0) +
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
    labs(title = "Map file and WGS assay metadata file") +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    theme(plot.margin = unit(c(0.5, 15, 1, 5), "cm"))

#Venn diagram showing overlap between map file and WGS assay file 
ggVennDiagram(list(`Map File "wgs_specimenID"` = qc_assay_map$'wgs_specimenID',
                   `Sample IDs` = sample_ids),
              label = "count", label_alpha = 0) +
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
    labs(title = "Map file and WGS assay metadata file") +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    theme(plot.margin = unit(c(0.5, 15, 1, 5), "cm"))


ggVennDiagram(
  list(
    `Map File`  = qc_assay_map$wgs_specimenID,
    `NYGC QC`   = wgs_qc_df$wgs_specimenID,
    `WGS Assay` = wgs_assay_df$specimenID
  ),
  label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  labs(title = "Overlap across the three files") +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = unit(c(0.5, 5, 1, 5), "cm"))




```

### lattice

```{r}
lattice::show.settings()
```

### base

```{r}
plot(pressure, col = thematic::thematic_get_option("accent"))
```





not_NYGC_samples <- qc_assay_map[qc_assay_map$filename.sample %in% setdiff(qc_assay_map$filename.sample, wgs_qc_df$Sample), 'wgs_specimenID']

hist(filtered_divco_wgs[filtered_divco_wgs$specimenID %in% not_NYGC_samples, "Percent.Sequence.Contamination"])


not_NYGC_samples_divco <- filtered_divco_wgs[filtered_divco_wgs$specimenID %in% not_NYGC_samples,]
not_NYGC_samples_contam <- filtered_divco_wgs[is.na(filtered_divco_wgs$Percent.Sequence.Contamination),]
NYGC_samples_contam_df <- filtered_divco_wgs[filtered_divco_wgs$Percent.Sequence.Contamination > 2,]

setdiff(NYGC_samples_contam, not_NYGC_samples_contam)
NYGC_samples_contam

qc_assay_map[grep('R6163593', qc_assay_map$final_specimenid), 'filename.sample']
grep('10447620-DLPFC', wgs_qc_df$Sample)
wgs_qc_df[427,]
qc_assay_map[grep('R2310801', qc_assay_map$final_specimenid), 'filename.sample']
grep("20917568-DLPFC", wgs_qc_df$Sample)
wgs_qc_df[47,]



wgs_assay_df[grepl('R2310801',wgs_assay_df$specimenID), ]




```{r}
library(dplyr)
library(tidyr)

# Metric columns: QC name -> assay name
metric_map <- c(
  "Total.Reads"               = "Total.Reads",
  "PF.Reads"                  = "PF.Reads",
  "X..PF.Reads"               = "Percent.PF.Reads",
  "PF.Aligned.Reads"          = "PF.Aligned.Reads",
  "X..PF.Aligned"             = "Percent.PF.Aligned",
  "PF.Aligned.Pairs"          = "PF.Aligned.Pairs",
  "X..PF.Aligned.Pairs"       = "Percent.PF.Aligned.Pairs",
  "Mean.Read.Length"          = "Mean.Read.Length",
  "Strand.Balance"            = "Strand.Balance",
  "Estimated.Library.Size"    = "Estimated.Library.Size",
  "Mean.Coverage"             = "Mean.Coverage",
  "X..Sequence.Contamination" = "Percent.Sequence.Contamination",
  "Median.Insert.Size"        = "Median.Insert.Size",
  "Mean.Insert.Size"          = "Mean.Insert.Size",
  "AT.Dropout"                = "AT.Dropout",
  "GC.Dropout"                = "GC.Dropout",
  "X..Total.Dup."             = "Percent.Total.Dup."
)

# Subset and rename each to a common metric name (use the assay name as canonical)
qc <- wgs_qc_df %>%
  filter(!is.na(wgs_specimenID)) %>%
  select(wgs_specimenID, all_of(names(metric_map))) %>%
  rename_with(~ metric_map[.x], .cols = names(metric_map)) %>%
  rename(specimenID = wgs_specimenID)

assay <- wgs_assay_df %>%
  select(specimenID, all_of(unname(metric_map)))

# Samples present in both
shared <- intersect(qc$specimenID, assay$specimenID)
cat("Shared samples:", length(shared), "\n")

# Long format, one row per (sample, metric) from each source, then compare
qc_long    <- qc %>% filter(specimenID %in% shared) %>%
  pivot_longer(-specimenID, names_to = "metric", values_to = "qc_value")
assay_long <- assay %>% filter(specimenID %in% shared) %>%
  pivot_longer(-specimenID, names_to = "metric", values_to = "assay_value")

tol <- 1e-6  # relative tolerance for floating-point metrics

cmp <- qc_long %>%
  inner_join(assay_long, by = c("specimenID", "metric")) %>%
  mutate(
    qc_value    = suppressWarnings(as.numeric(qc_value)),
    assay_value = suppressWarnings(as.numeric(assay_value)),
    abs_diff    = abs(qc_value - assay_value),
    denom       = pmax(abs(qc_value), abs(assay_value)),
    rel_diff    = ifelse(denom == 0, 0, abs_diff / denom),   # both-zero -> 0, not NaN
    status = case_when(
      is.na(qc_value) & is.na(assay_value)              ~ "both_missing",
      is.na(qc_value) | is.na(assay_value)              ~ "one_missing",
      rel_diff <= tol                                    ~ "match",
      TRUE                                               ~ "mismatch"
    )
  )

cmp %>%
  count(metric, status) %>%
  pivot_wider(names_from = status, values_from = n, values_fill = 0) %>%
  print(n = Inf)

# the genuinely interesting rows
cmp %>% filter(status == "mismatch") %>% arrange(desc(rel_diff)) %>% print(n = Inf)

# samples missing on one side, for follow-up
cmp %>% filter(status == "one_missing") %>% distinct(specimenID) %>% print(n = Inf)


fix_ids <- c("R2310801", "R6163593")

# QC file is keyed on wgs_specimenID; assay on specimenID
for (id in fix_ids) {
  qc_row <- wgs_qc_df[wgs_qc_df$wgs_specimenID == id & !is.na(wgs_qc_df$wgs_specimenID), ]
  if (nrow(qc_row) != 1) {
    warning("Expected exactly 1 QC row for ", id, " but found ", nrow(qc_row), " — skipping")
    next
  }
  assay_idx <- which(wgs_assay_df$specimenID == id)
  if (length(assay_idx) != 1) {
    warning("Expected exactly 1 assay row for ", id, " but found ", length(assay_idx), " — skipping")
    next
  }
  # metric_map: names = QC columns, values = assay columns
  for (qc_col in names(metric_map)) {
    assay_col <- metric_map[[qc_col]]
    wgs_assay_df[assay_idx, assay_col] <- qc_row[[qc_col]]
  }
}

wgs_assay_df[wgs_assay_df$specimenID %in% fix_ids,
             c("specimenID", unname(metric_map))]


write.table(wgs_assay_df, '/home/ec2-user/data/processed/metadata/DivCO/DivCo_assay_WGS_metadata.csv', sep = ',', row.names = F)

```

wgs_assay_df <- wgs_assay_df %>% mutate('Batch' = case_when(wgs_assay_df$specimenID %in% wgs_qc_df$wgs_specimenID ~ 'Batch 1',
                                                          TRUE ~ 'Batch 2'))

table(wgs_assay_df$Batch)                                                          


library(ggplot2)
wgs_assay_df$fill_group <- ifelse(
  wgs_assay_df$Percent.Sequence.Contamination > 2,
   wgs_assay_df$Batch,
  "Below cutoff"
)

fill_vals <- c(
  "Below cutoff" = "grey80",
  "Batch 1"      = "#2166AC",
  "Batch 2"      = "#B2182B"
)

contam_hist <- ggplot(wgs_assay_df, aes(x = Percent.Sequence.Contamination, fill = fill_group)) +
  geom_histogram(binwidth = 0.1, colour = "grey30", linewidth = 0.1) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = fill_vals) +
  annotate("text", x = 2.1, y = 50, label = "Contamination cutoff (> 2%)",
           colour = "black", hjust = 0, size = 3.5, fontface = "italic") +
  labs(title = "Sequence contamination",
       subtitle = "Contaminated samples (> 2%) colored by batch; others grey",
       x = "Percent Sequence Contamination", y = "Count",
       fill = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")


  ggsave(file.path('/home/ec2-user/AMP-AD_genetic_PCs/Results/Filtering/Sequence_contamination', "Percent_contamination_hist.png"),
      contam_hist, width = 11, height = 8, dpi = 150)






contam_samples <- wgs_assay_df[wgs_assay_df$Percent.Sequence.Contamination > 2, ]

jitter <- ggplot(contam_samples, aes(x = Percent.Sequence.Contamination, y = 0)) +
  geom_jitter(height = 0.15, size = 3, alpha = 0.8, colour = "#B2182B") +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "grey40", linewidth = 0.7) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  annotate("text", x = 2.60, y = 0.25, label = "3%", colour = "black", size = 6, fontface = "italic") +
  annotate("text", x = 4.60, y = 0.25, label = "5%", colour = "black", size = 6, fontface = "italic") +
  labs(title = "Samples above contamination cutoff (> 2%, all Batch 2)",
       x = "Percent Sequence Contamination", y = NULL) +
  theme_bw(base_size = 12) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")


  ggsave(file.path('/home/ec2-user/AMP-AD_genetic_PCs/Results/Filtering/Sequence_contamination', "Over_two_percent_contam_jitter.png"),
      jitter , width = 11, height = 8, dpi = 150)

ggplot(wgs_assay_df, aes(x = Percent.Sequence.Contamination, fill = factor(Batch))) +
  geom_histogram(binwidth = 0.25, colour = "grey30") +
  scale_fill_manual(values = c("1" = "#2166AC", "2" = "#B2182B")) +
  coord_cartesian(xlim = c(2, NA)) +
  labs(x = "Percent Sequence Contamination", y = "Count", fill = "Batch") +
  theme_bw(base_size = 12)


  dim(contam_samples[contam_samples$Percent.Sequence.Contamination <5, ])
